---
title: "data_generation"
author: "Haolin Zhong"
date: "2022/2/8"
output: github_document
---

```{r}
library(tidyverse)
library(caret)
library(survival)
```


## Culmulative Distribution of Time t

$$S(t) = \displaystyle \int_t^{\infty} f(t)dt = 1 - F(t)$$

$$h(t) = \frac{f(t)}{S(t)} = \frac{F'(t)}{1 - F(t)}$$

$$F(t) = \displaystyle 1 - \text{exp}\left(-\int_0^t h(s)ds\right) = 1 - \text{exp} \left[ -(\int_0^t h_0(s)ds)  \cdot  \text{exp}(\beta^T X) \right]$$

  - exponential proportional-hazards model: 

$$F(t) = 1 - \text{exp}\left[- \lambda t \cdot \exp(\beta^T X) \right] \quad \implies \quad F^{-1}(U) = - \frac{\ln(U)}{\lambda \exp(\beta^TX)}$$
  
  - Weibull proportional-hazards model: 

$$F(t) = 1 - \text{exp}\left[- \lambda t^{\gamma} \cdot \exp(\beta^T X) \right] \quad \implies \quad F^{-1}(U) = \left( - \frac{\ln(U)}{\lambda \exp(\beta^TX)}\right) ^{\frac 1 {\gamma}}$$
  
  - Geompartz proportional-hazards model:
  
$$
F(t) = 1 - \text{exp}\left[- \lambda t^{\gamma} \cdot \exp(\beta^T X) \right] \quad \implies \quad F^{-1}(U) = \frac 1 {\alpha} \ln{\left(1 - \frac{\alpha\ln U}{\lambda \exp{\beta^T X}}\right)}
$$
  
```{r}
cox = function(u, x, lambda, alpha, beta){
    time = (1/alpha) * log(1 - (alpha * log(u) / (lambda * exp(beta * x))))
  }
```



## Data Generation

- generate mixed random t under given sample size `n` and mixing proportion `p`:
  
  - when `p = 1`, use pure exponential distribution
  
  - when `p = 0`, use pure weibull distribution
  
  - This function use lambda = 0.5, gamma = 1.5 as default
  
  - In later analysis we will always use `b1 = 0.5` as the treatment effect.
  
```{r}

simdat = function(n, p, lambda = 0.5, gamma = 1.5, eff = list()) {
  # randomly assign group
  x1 = rbinom(n, 1, 0.5)
  
  u = runif(n)
  useExp = runif(n) < p 
  # generating mixed data
  t =  useExp * (-log(u) / (lambda * exp(eff$b1 * x1))) + (1 - useExp) * (-log(u) / (lambda * exp(eff$b1 * x1)))^(1 / gamma)
  
  # two features
  # x2 = rbinom(n, 1, 0.5)
  # t =  useExp * (-log(u) / (lambda * exp(eff$b1 * x1))) + (1 - useExp) * (-log(u) / (lambda * exp(eff$b2 * x2)))^(1 / gamma)
  
  t[t < 1/365] = 1/365
  t[t == 1 / 365] = t[t == 1 / 365] + rnorm(length(t[t == 1 / 365]), 0, 1e-4)
  t = abs(t)
  
  e = as.numeric(t < 5)
  t = pmin(t, 5)
  
  name = paste("n =", n, ", p =", p, ", eff =", eff[[1]])
  
  return(tibble(name = name, time = t, event = e, x1 = x1))
  # two features
  # return(tibble(name = name, time = t, event = e, x1 = x1, x2 = x2))
}
```


- test:

```{r}
data = simdat(500, 0, eff = list(b1 = 0.5))

data %>% 
  ggplot() +
  geom_density(aes(x = time))
```

## fit function

```{r}
fit_exp = function(df) {
  fit.exponential = survreg(Surv(time, event) ~ x1, dist = "exponential", data = df)
  return(as.numeric(-fit.exponential$coefficients[-1]))
}

fit_weibull = function(df) {
  fit.weibull <- survreg(Surv(time, event) ~ x1, dist = "weibull", data = df)
  return(as.numeric(-fit.weibull$coefficients[-1] / fit.weibull$scale))
}

fit_cox = function(df) {
  fit.cox <- coxph(Surv(time, event) ~ x1, data = df)
  return(as.numeric(fit.cox$coefficients))
}
```



## The distribution of estimated beta

- visualization:

```{r, fig.width=10}
param_grid = expand.grid(p = seq(0, 1, by = 0.2), n = c(20, 40, 60, 80, 100, 200, 400), rep = 1:100)

sim_dat = param_grid %>% 
  mutate(
    data = map2(n, p, ~simdat(n = .x, p = .y, eff = list(b1 = 0.5)))
  ) %>% 
  mutate(
    b_exp = map_dbl(data, fit_exp),
    b_weibull = map_dbl(data, fit_weibull),
    b_cox = map_dbl(data, fit_cox)
  )

sim_dat %>% 
  ggplot() +
  geom_density(aes(x = b_exp, color = "darkred"), alpha = 0.6) +
  geom_density(aes(x = b_weibull, color = "orange" ), size = 1.2) +
  geom_density(aes(x = b_cox, color = "blue"), alpha = 0.3) +
  geom_vline(xintercept = 0.5) +
  facet_grid(n~p) +
  xlab("estimated beta") +
  scale_colour_manual(name = 'fit methods', 
        values =c('blue'='blue','darkred'='darkred', 'orange' = 'orange'), labels = c('cox','exp', 'weibull'))
```


- metrics: 
  
  - MSE may reflect accuracy: (the trend is indeed very obvious)
  
```{r}
mse = sim_dat %>% 
  select(-data, -rep) %>%
  mutate(
    b_exp = (b_exp - 0.5)^2,
    b_weibull = (b_weibull - 0.5)^2,
    b_cox = (b_cox - 0.5)^2
  ) %>% 
  group_by(n, p) %>% 
  summarize(
    exp_mse = mean(b_exp),
    weibull_mse = mean(b_weibull),
    cox_mse = mean(b_cox)
  ) %>% 
  pivot_longer("exp_mse":"cox_mse", names_to = "fit_method", values_to = "mse") %>% 
  mutate(fit_method = as.factor(fit_method)) 
  
  
ggplot(mse, aes(x = n, y = mse, color = fit_method)) +
  geom_point() +
  geom_line() +
  scale_y_log10() +
  facet_grid("p")
```


```{r}
sim_dat %>% 
  select(-data, -rep) %>% 
  pivot_longer("b_exp":"b_cox", names_to = "fit_method", values_to = "beta") %>% 
  ggplot(aes(x = interaction(n), y = beta, color = fit_method)) +
  geom_boxplot() +
  facet_grid(fit_method~ p)
```



